---
title: "MUVR rdCV-RF Data Analysis and Validation based on BMI, WC and %BF categories"
author: 
- "Cristina Artero Martínez (Main author)"
- "Enrique Almanza Aguilera (Supervision & Edition)"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    higlight: kate
    code_folding: hide
    toc_depth: 4
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
---

<style type = "text/css">
h1.title{ /* Title */
  font-size: 40px
}
h1{ /* Header 1 */
  font-size: 24px
}
h2{ /* Header 2 */
  font-size: 20px
}
h3{ /* Header 3 */
  font-size: 18px
}
h4{ /* Header 4 */
  font-size: 15px
}
h5{ /* Header 5 */
  font-size: 12px
}
</style>


```{r startpoint, include = FALSE,echo = FALSE,message=FALSE}
startpoint=Sys.time()
```

The following report summarizes the models obtained with rdcv-RF for the different proteomics panels based on the BMI, WC and %BF categories. Note that models were sex-stratified (i.e., performed on all individuals, only females and only males).


# Setting and loading packages + setting directories
Packages and directories need to be prepared in order to have easier ways to operate and access data.
```{r Loading packages, message=FALSE, warning = FALSE,echo = FALSE}
# With the following list, packages are automatically loaded in the next instruction
packages=c("doParallel","Rdisop","kableExtra","magrittr","rmarkdown","BioMark","MUVR","impute","missForest","plyr","ropls","parallel","factoextra","dplyr","BiocParallel")

#install.pack=packages[!packages %in% installed.packages()]
#for(i in install.pack) install.packages(i,dependencies=TRUE)

# With this instruction, packages are loaded properly
invisible(lapply(packages, library, character.only = TRUE))

# Preparing the directories
# This directory represents the main directory (it contains the other directories, defined below)
dir.datasets = "/home/crisam/crisam/datasets/"
# This directory contains proteomics data of participants (here there are datasets corresponding to proteomics panels)
dir.proteo = "/home/crisam/crisam/datasets/proteomics/"
# This directory contains clinical data of participants (here there is stored the clinical examination dataset)
dir.clin = "/home/crisam/crisam/datasets/clinical/"
```


# Loading datasets
Different datasets need to be loaded in order to model data. Below, in comments, there are specifications about them.
```{r}
# Loading the dataset with the population of interest
delta.bwa_df = readRDS(file = paste0(dir.clin, "delta.bwa.omics.rds")) # this dataset corresponds to the Rdata object containing the dataset based on the population of interest
prot = delta.bwa_df$delta.bwa.prot # this is the dataset with the population of interest

# Loading the proteomics panels
CVD2 = read.csv(paste0(dir.proteo, "CVD2.csv")) # this is Cardiovascular II panel
CVD3 = read.csv(paste0(dir.proteo, "CVD3.csv")) # this is Cardiovascular III panel
metabolism = read.csv(paste0(dir.proteo, "Metabolism.csv")) # this is cardiomaetabolic panel

# Loading the clinical examination dataset (the one containing waist circumferences (WC) and fat mass measures)
clin = read.csv(paste0(dir.clin, "visitclin.csv"))
```


# Predefining some variables
Some variables need to be reformatted or included. Specifically, those that represent a factor or a date variable. This is to ensure the proper manipulation of data for posterior analyses.
```{r}
# Reformatting
# Sex must be a factor variable
prot$sex = as.factor(prot$sex)
# The cohort is also a factor variable and needs to be formatted as it
prot$project = as.factor(prot$project)
# The visitdate variable needs to be formatted as a date variable instead of a string
prot$visitdate = as.Date(prot$visitdate, "%Y-%m-%d")

# Including
# WC must be incorporated to the dataset of the population of interest
prot = merge(prot, clin[,c("SIMPKEY", "Waist")])
# %BF must be computed and included to the dataset of the population of interest
clin$X100_fett_total = (clin$fett_total/1000)/clin$weight * 100
prot = merge(prot, clin[,c("SIMPKEY", "X100_fett_total")])
```


# Preprocessing the proteomics panels
Once all datasets have been loaded, we must perform a few preprocessing steps to ensure that:
* the study population is only being considered.
* the variable of interest is being included in the dataset (i.e., BMI, WC, %BF).
* categories are classified based on their variables' particular thresholds.
* individuals or proteins with a high number of missings are removed.
```{r}
# The function takes 2 arguments: a dataset (one of the proteomics panels) that will be preprocessed based on the conditions specified above, and a variable name that will determine the variable which will be the main response variable from the dataset
preprocess_dataset = function(dataset, variable){
  # The panels do not have the project updated and may need it up to date
  dataset$project = NULL
  # The script focuses on 3 main variables of exposure: BMI, WC and %BF. This means that they need to be included to each proteomic panel
  dataset = merge(dataset, prot[, c("SIMPKEY", "project", "bmi.clin", "Waist", "X100_fett_total")], by = "SIMPKEY")
  # In the case of working with a certain variable, the population of interest will contemplate several participants or not. Therefore, for each variable, only individuals with information (no missing values) will be considered
  if (variable == "bmi.clin"){
    dataset = dataset[complete.cases(dataset$bmi.clin),]
    dataset = dataset[dataset$bmi.clin >= 18.5 & dataset$bmi.clin < 40,]
    # For BMI tests two main comparisons were performed:
    # - Normal weight vs Overweight vs Obesity  
    dataset$bmi.category = as.factor(ifelse(dataset$bmi.clin < 25, "normal weight", ifelse(dataset$bmi.clin < 30, "overweight", "obesity")))
    # - Normal weight vs Overweight + Obesity 
  dataset$bmi.category2 = as.factor(ifelse(dataset$bmi.clin < 25, "normal weight", "overweight + obesity"))
  }
  else if (variable == "Waist"){
    dataset = dataset[complete.cases(dataset$Waist),]
    # For WC tests three main comparisons were performed:
    # - No metabolic risk vs Increased metabolic risk vs Substantially increased metabolic risk  
    dataset$wc.category = as.factor(ifelse(dataset$Waist < 80 & dataset$sex == "F", "no metabolic risk", ifelse(dataset$Waist < 88 & dataset$sex == "F", "increased metabolic risk", ifelse(dataset$Waist > 88 & dataset$sex == "F", "substantially increased metabolic risk", ifelse(dataset$Waist < 94 & dataset$sex == "M", "no metabolic risk", ifelse(dataset$Waist < 102, "increased metabolic risk", "substantially increased metabolic risk"))))))
    # - No metabolic risk vs Increased metabolic risk + Substantially increased metabolic risk  
  dataset$wc.category2 =  as.factor(ifelse(dataset$Waist < 80 & dataset$sex == "F", "no metabolic risk", ifelse(dataset$Waist < 88 & dataset$sex == "F", "metabolic risk", ifelse(dataset$Waist < 94 & dataset$sex == "M", "no metabolic risk", "metabolic risk"))))
  # - No metabolic risk vs Substantially increased metabolic risk  
  dataset$wc.category3 = as.factor(ifelse(dataset$Waist < 80 & dataset$sex == "F", "no metabolic risk", ifelse(dataset$Waist < 88 & dataset$sex == "F", "-", ifelse(dataset$Waist > 88 & dataset$sex == "F", "substantially increased metabolic risk", ifelse(dataset$Waist < 94 & dataset$sex == "M", "no metabolic risk", ifelse(dataset$Waist < 102, "-", "substantially increased metabolic risk"))))))
  }
  else{
    dataset = dataset[complete.cases(dataset$X100_fett_total),]
    # For %BF tests the comparisons was normal weight vs obesity. Note that although men are contemplated, men in the study did not have DEXA information, so in the end they will not be included in the model 
    dataset$fett.category = as.factor(ifelse(dataset$X100_fett_total < 35 & dataset$sex == "F", "normal weight", ifelse(dataset$X100_fett_total > 35 & dataset$sex == "F", "obesity", ifelse(dataset$X100_fett_total < 25 & dataset$sex == "M", "normal weight", "obesity"))))
  }
  # As explained before, proteins or individuals with more or equal than 50% of missings must be rejected.
  dataset = dataset[rowSums(is.na(dataset)) <= 92/2,]
  dataset = dataset[,colSums(is.na(dataset)) <= nrow(dataset)/2]
  
  return(dataset)
}
```

Proteomics panels are preprocessed based on the conditions previously mentioned. As depending on the variable, the number of observations will vary, it is safer to store the preprocessed panels in different objects.
```{r}
# Here, the panels are preprocessed based on BMI
CVD2.bmi = preprocess_dataset(CVD2, "bmi.clin")
CVD3.bmi = preprocess_dataset(CVD3, "bmi.clin")
metabolism.bmi = preprocess_dataset(metabolism, "bmi.clin")
# Here, the panels are preprocessed based on WC (the number of individuals is lower)
CVD2.wc = preprocess_dataset(CVD2, "Waist")
CVD3.wc = preprocess_dataset(CVD3, "Waist")
metabolism.wc = preprocess_dataset(metabolism, "Waist")
# Here, the panels are preprocessed based on %BF (the number of individuals is even lower)
CVD2.bf = preprocess_dataset(CVD2, "bf")
CVD3.bf = preprocess_dataset(CVD3, "bf")
metabolism.bf = preprocess_dataset(metabolism, "bf")
```


# Misclassifications in categories
Knowing the misclassification rate (MISS) or the area under the curve (AUC) of the model is not enough. Additional information such as the misclassification rate for each group is important to understand if the model is more biased to one class or not. The following function will be useful to determine the number of misclassifications per class.
```{r}
# The function takes the original Y-vector and the predicted one
MISS_per_classifications = function(ydata, model_ydata){
  # By creating a table, we compare categories, therefore, we can check how many observations where correctly predicted by category
  mat = unclass(table(ydata, model_ydata))
  # As we are interested only on misclassifcations, the diagonal (correctly predicted observations) are not interesting
  diag(mat) = 0
  return(rowSums(mat))
}
```


# Sex-stratification
One of the main objectives in the study was to understand anthropometric measures on 2 Swedish cohorts, which are composed of different sex. To perform that, this function will be useful to create datasets based on the sex of participants.
```{r}
# The function takes 2 arguments: a dataset (one of the proteomics panels) that will be sex-stratified, and the sex which will define the stratification
subset_by_sex = function(dataset, sex){
  # In the case of specifying female,
  if (sex == "F"){
    # the dataset will only include female indivduals
    subset = dataset[dataset$sex == "F",]
  }
  # If the specification refers to male,
  else{
    # the dataset will be subsetted to male individuals
    subset = dataset[dataset$sex == "M",]
  }
  
  return(subset)
}
```


# Modeling
The model parameters are the same for all models, the only difference is the used response variable and the specified sex-stratification. A simpler way to model could be by defining a function to automatically do it. As a reminder, model parameters where:
* the proteome panel/dataset (X)
* a response vector corresponding to the exposure variable (Y)
* a vector of unique variable identifiers (ids)
* the number of repetitions (nRep = 10)
* the number of outer cross-validation segments (nOuter = 8)
* the variable ratio (proportion of variables kept per iteration, varRatio = 0.7)
```{r}
model.rdcv = function(dataset, response, sex){
  # This condition is to check whether we have specified to run the model with a certain sex group or not.
  if (!missing(sex)){
    dataset = subset_by_sex(dataset, sex)
  }
  
   # First, we must prepare the X matrix, the Y vector and the IDS vector. We must also impute values in case they are NA (as the algorithm cannot classify missing values). A scaling step is recommended.
  xdata = dataset[, grep("^OID", colnames(dataset))]
  xdata = impute.knn(t(xdata), k=10, rowmax = 60, colmax = 60, maxp = 5000)
  xdata = t(xdata$data)
  ydata = dataset[, grep(response, colnames(dataset))]
  kable(table(ydata), caption = "Distribution of categories", col.names = c("Categories", "Frequency"))
  idx = dataset$SIMPKEY
  
  # The following parameters refer to parameters from the modeling function. Notice that the number of cores is not equal tot the total. That is to allow external computations and avoid a risk of a collapse. 
  ncores=detectCores()-1
  nRep=ncores
  nOuter=8
  nInner=nOuter-1
  varRatio=0.70
  methParam.RF=customParams(method = 'RF',ntreeIn = 150,ntreeOut = 300)
  
  # An important step is to parallelize. This is to optimize the execution of the code. If not, the model would take hours to finish. 
  cl=makeCluster(ncores)
  registerDoParallel(cl)
  
  # The model is performed based on the parameters from above. Check that after it, the parallelization must stop.
  model = MUVR(X = xdata, Y = ydata, ID = idx, nRep = nRep, nOuter = nOuter, nInner = nInner, varRatio = varRatio, method = 'RF', fitness = 'MISS', methParam = methParam.RF)
  
  # After having performed the model, parallelization is no longer needed
  stopCluster(cl)
  
  print("Number of misclasifications")
  model$miss
  modelFit = model$miss[2]
  print(paste0("Error rate (%): ", round(modelFit/nrow(xdata)*100,2), "%"))
  
  print("Plot Misclasifications by category")
  plotMV(model,model = "mid")
  
  
  print("AUC classifications")
  model$auc
  
  print("Number VIP Features")
  mod.vips = getVIP(MVObj = model, model = "mid")
  nrow(mod.vips)
  
  print("Misclassifcations per classification groups")
  mpc = MISS_per_classifications(ydata, model$yClass$mid)
  print(mpc)
  print(paste0(round(mpc/modelFit*100,2), "%"))
  
  return(model)
}
```


# Individual panels
BMI was explored not only on all proteomics panels combined but also on individual panels. Therefore, there is the code to explore each panel on each sex-stratification for BMI.
```{r rdcv-RF for individual panels for BMI categories, warning = FALSE, echo = FALSE, message=FALSE}
model1 = model.rdcv(CVD2.bmi, "bmi.category$")
model2 = model.rdcv(CVD2.bmi, "bmi.category2")
model3 = model.rdcv(CVD2.bmi, "bmi.category$", "F")
model4 = model.rdcv(CVD2.bmi, "bmi.category2", "F")
model5 = model.rdcv(CVD2.bmi, "bmi.category$", "M")
model6 = model.rdcv(CVD2.bmi, "bmi.category2", "M")
model7 = model.rdcv(CVD3.bmi, "bmi.category$")
model8 = model.rdcv(CVD3.bmi, "bmi.category2")
model9 = model.rdcv(CVD3.bmi, "bmi.category$", "F")
model10 = model.rdcv(CVD3.bmi, "bmi.category2", "F")
model11 = model.rdcv(CVD3.bmi, "bmi.category$", "M")
model12 = model.rdcv(CVD3.bmi, "bmi.category2", "M")
model13 = model.rdcv(metabolism.bmi, "bmi.category$")
model14 = model.rdcv(metabolism.bmi, "bmi.category2")
model15 = model.rdcv(metabolism.bmi, "bmi.category$", "F")
model16 = model.rdcv(metabolism.bmi, "bmi.category2", "F")
model17 = model.rdcv(metabolism.bmi, "bmi.category$", "M")
model18 = model.rdcv(metabolism.bmi, "bmi.category2", "M")
```


# All panels
The following code will be used to combine all panels into a single one. As panels were preprocessed differently for each variable of exposure (BMI, WC and %BF), the combination of panels appears in three different ways.
```{r}
# BMI
all.bmi = merge(CVD2.bmi[, c(1,grep("^OID", colnames(CVD2.bmi)))], CVD3.bmi[, c(1,grep("^OID", colnames(CVD3.bmi)))], by = "SIMPKEY")
all.bmi = merge(all.bmi, metabolism.bmi[, grep("SIMPKEY|sex|bmi.category|^OID", colnames(metabolism.bmi))], by = "SIMPKEY")
# WC
all.wc = merge(CVD2.wc[, c(1,grep("^OID", colnames(CVD2.wc)))], CVD3.wc[, c(1,grep("^OID", colnames(CVD3.wc)))], by = "SIMPKEY")
all.wc = merge(all.wc, metabolism.wc[, grep("SIMPKEY|sex|wc.category|^OID", colnames(metabolism.wc))], by = "SIMPKEY")
# %BF
all.bf = merge(CVD2.bf[, c(1,grep("^OID", colnames(CVD2.bf)))], CVD3.bf[, c(1,grep("^OID", colnames(CVD3.bf)))], by = "SIMPKEY")
all.bf = merge(all.bf, metabolism.bf[, grep("SIMPKEY|sex|fett.category|^OID", colnames(metabolism.bf))], by = "SIMPKEY")
```

## Modeling all panels
Below, there are the models performed on BMI, WC and %BF. Note that BMI and WC models incorporate sex-stratifications whereas %BF no (as SMC-C Uppsala individuals were the only participants containing information about fat mass measurements).
```{r rdcv-RF for all panels for BMI and WC and %BF, warning = FALSE, echo = FALSE, message=FALSE}
model19 = model.rdcv(all.bmi, "bmi.category$")
model20 = model.rdcv(all.bmi, "bmi.category2")
model21 = model.rdcv(all.bmi, "bmi.category$", "F")
model22 = model.rdcv(all.bmi, "bmi.category2", "F")
model23 = model.rdcv(all.bmi, "bmi.category$", "M")
model24 = model.rdcv(all.bmi, "bmi.category2", "M")

model25 = model.rdcv(all.wc, "wc.category$")
model26 = model.rdcv(all.wc, "wc.category2")
model27 = model.rdcv(all.wc, "wc.category3")
model28 = model.rdcv(all.wc, "wc.category$", "F")
model29 = model.rdcv(all.wc, "wc.category2", "F")
model30 = model.rdcv(all.wc, "wc.category3", "F")
model31 = model.rdcv(all.wc, "wc.category$", "M")
model32 = model.rdcv(all.wc, "wc.category2", "M")
model33 = model.rdcv(all.wc, "wc.category3", "M")

model34 = model.rdcv(all.bf, "fett.category")
```


# Saving the models
The completion of all models may take several hours. This supposes a lot of time and resources. Therefore, results from models should be stored in Rdata objects to easily retrieve their information. This step is to save the models in order to later work with them.
```{r}
models = list("model1" = model1, "model2" = model2, 
              "model3" = model3, "model4" = model4, 
              "model5" = model5, "model6" = model6,
              "model7" = model7, "model8" = model8, 
              "model9" = model9, "model10" = model10, 
              "model11" = model11, "model12" = model12,
              "model13" = model13, "model14" = model14,
              "model15" = model15, "model16" = model16, 
              "model17" = model17, "model18" = model18,
              "model19" = model19, "model20" = model20, 
              "model21" = model21, "model22" = model22, 
              "model23" = model23, "model24" = model24,
              "model25" = model25, "model26" = model26, 
              "model27" = model27, "model28" = model28,
              "model29" = model29, "model30" = model30, 
              "model31" = model31, "model32" = model32, 
              "model33" = model33, "model34" = model34)

saveRDS(models, file = paste0(dir.datasets,"models.RF.rds"))
```


# Validation
Once models were obtained, significant ones needed to be validated. To do that, permutation testing was used as seen below. Check that the parameters used are the same as in standard modeling.
```{r rdcv-RF permutations, warning = FALSE, echo = FALSE, message=FALSE}
# The following function takes as arguments the model to validate, its X-matrix, its Y-vector, its ids and its number of misclassifications
validation = function(model, xdata, ydata, idx, MISS){
  # As before, parallelization is required, so we must first set the number of cores (again: all cores minus one)
  ncores=detectCores()-1
  # After setting the parameter, parallelization starts to optimize the code
  cl=makeCluster(ncores)
  registerDoParallel(cl)
  
  # The number of permutations must be a sufficient one. 20 are more than enough
  for (p in 1:20) {
    cat('\nPermutation',p,'of',20)
    # The Y-vector is sampled, so labels are shuffled
    YPerm = sample(ydata)
    # The model is performed on the modified Y-vector
    perm = MUVR(X = xdata, Y = YPerm, ID = idx, nRep = 10, nOuter = 8, nInner = 7, varRatio = 0.7, method = 'RF', fitness = 'MISS', methParam = methParam.RF)
    # The only important parameter is the misclassification from the permutation, so it must be stored
    permFit[p]=perm$miss[2]
  }

  print("Permutation Test")
  # Once all misclassifications are stored a parametric comparison is performed to evaluate the quality of parameters 
  pPerm(actual = MISS, h0 = permFit, type = "t")
  # This plot will help to compare the model's number of misclassifications to the ones from permutations from the parametric test
  plotPerm(actual = MISS, h0 = permFit, type="t") 
  # Here, a non-parametric comparison is performed to also evaluate the quality of parameters
  pPerm(actual = MISS, h0 = permFit, type = "non")
  # This plot will help to compare the model's number of misclassifications to the ones from permutations from the non-parametric test
  plotPerm(actual = MISS, h0 = permFit, type="non") 
  
  # Finally, parallelization ends as it is no longer required
  stopCluster(cl)
}
```

After having defined the function, validation can start for selected models.
```{r}
test8.2 = validation(model20, model20$inData$X, model20$inData$Y, model20$inData$ID, model20$miss[2]) # BMI on F & M
test8.2F = validation(model22, model22$inData$X, model22$inData$Y, model22$inData$ID, model22$miss[2]) # BMI on F
test12.3 = validation(model27, model27$inData$X, model27$inData$Y, model27$inData$ID, model27$miss[2]) # WC on F & M
test12.3F = validation(model30, model30$inData$X, model30$inData$Y, model30$inData$ID, model30$miss[2]) # WC on F
test13.1 = validation(model34, model34$inData$X, model34$inData$Y, model34$inData$ID, model34$miss[2]) # %BF (on F)
```


# Saving the data
After having performed models and their respective validation tests, one way to ensure that results are not lost is by saving the Rdata into a file.
```{r}
save.image(file=paste0("rdCV.RF.validation.RData"))
```


